The yaq Project: \\ Standardized Software Enabling Flexible Instrumentation

### 2023-02-06 paper-as-submitted.tex
diff

Modern instrumentation development often involves incorporation of many dissimilar hardware peripherals into a single unified instrument.
Increasing availability of modular hardware has brought greater instrument complexity to small research groups.
This complexity stretches the capability of traditional, monolithic orchestration software.
In many cases, a lack of software flexibility leads creative researchers to feel frustrated, unable to perform experiments they envision.
Herein we describe \yaq{}, a software project defining a new standardized way of communicating with diverse hardware peripherals.
\yaq{} encourages a highly modular approach to experimental software development which is well suited to address the experimental flexibility needs of complex instruments.
\yaq{} is designed to overcome hardware communication barriers that challenge typical experimental software.
A large number of hardware peripherals are already supported, with tooling available to expand support.
The \yaq{} standard enables collaboration among multiple research groups, increasing code quality while lowering development effort.

@@

We have created a new network-based communication standard for scientific instrumentation, \yaq{}.
This standard borrows the most important ideas from established projects used by large user facilities while retaining the simplicity appropriate for small research labs.
We have built this standard to be self-describing, portable, and reusable wherever possible.
Via the \yaq{} interface, we provide easy-to-use hardware support for scientists to use with their choice of experiment orchestration software.

@@

In \yaq{}, communication between daemons and clients is performed over TCP/IP using Apache Avro RPC \cite{AvroSpecification}.
Avro provides an agreed upon standard for efficient serialization of data and method calls from a remote (client) process.
Practically, the \yaq{} interface looks like a collection of methods or functions, which Avro calls ``messages''.
Each message has defined input parameters and output return types.
A sensor might implement a message called ``\texttt{get\_measured}'' which takes no parameters and returns a dictionary mapping channel names to numeric or array data.
A motor would implement a pair of messages for setting and reading back the motor position: ``\texttt{set\_position(float position)} $\rightarrow$ \texttt{null}'' and ``\texttt{get\_position()} $\rightarrow$ \texttt{float}''.
Messages make up the lowest level functionality of the \yaq{} interface.
Each individual communication between client and daemon involves one message being requested by the client and the response returned from the daemon.
If multiple messages are sent simultaneously, the daemon processes one message at a time.
The protocol is preemptive: in the case of a conflict, the daemon will obey the last message to arrive.
Each daemon supports a collection of messages for its unique functionality, called the ``protocol''.
When a client first connects to a daemon over TCP/IP, the daemon provides a complete description of its own protocol.
We therefore refer to the daemons as ``self-describing''.

\yaq{} introduces a concept called ``traits'' which are collections of related messages that are shared among multiple protocols.
Motors implement the ``\texttt{has-position}'' trait, which defines ``\texttt{set\_position}'', ``\texttt{get\_position}'', and ``\texttt{get\_units}''.
Sensors would implement the ``\texttt{is-sensor}'' trait which defines ``\texttt{get\_measured}'', ``\texttt{get\_channel\_names}'', and ``\texttt{get\_channel\_units}''.
Protocols which implement a trait must support all of the messages from the trait.
Traits are compositional: a given protocol may implement several traits at once.
For example, a protocol representing a monochromator with several gratings might implement both ``\texttt{has-position}`` and ``\texttt{has-turret}``, where the latter defines special messages for choosing which grating is used.
Importantly, specific protocols can also implement arbitrary additional messages that are not defined by any trait.

We have carefully defined \yaq{} traits to maximize interoperability while not excluding hardware with unusual features.
New traits can be discussed by the community through the yaq enhancement proposal system.
It's crucial that multiple hardware interface examples be considered when defining traits.
Configuration that is unique to a single interface is best handled through protocol-specific messages.
For example, our protocol for interfacing with data acquisition cards from Gage provides a message for setting the segment count.
Unique messages are avaliable through script-based and graphical clients, but usage of such messages in scripts naturally limits portability.

We now describe how the \yaq{} architecture addresses the three hardware interface barriers described at the beginning of this section.
The first hardware interface barrier: Multiple incompatible interfaces are used to communicate with each component of the system. % consider making this sentence (and the next paragraph starts) subsection headings
\yaq{} provides a unified TCP/IP interface to all hardware peripherals based on the well-described Avro RPC protocol.
The trait system was introduced in pursuit of our primary goal of easing the client development process.
Clients can trust that protocols that implement a given trait will behave in similar ways.
The standardized \yaq{} interface presents the same set of interactions for client-side scientific code, simplifying the experience of using hardware.

@@

The third hardware interface barrier: Some hardware interfaces are blocking or slow.
Experimental orchestration often means doing several things at once, \textit{e.g.} simultaneously moving several motors.
To accomplish this task performantly, monolithic orchestration software often necessitates separate threads for each component hardware interface, a fragile pattern in which small mistakes become both critical and elusive errors.
When writing orchestration software using \yaq{}, slow hardware interfaces are replaced by fast Avro-RPC.
In our experience, the \yaq{} interface is responsive enough to give good performace when orchestrating tens of motors using a singly-threaded client.
There is a limit where using TCP becomes a bottleneck for large resonses, such as those from cameras or other high throughput sensors.
In our experience, this limit is approximately 1 megapixel.
For larger sensors, a pure \yaq{} approach may not be appropriate.
There is a longer discussion of timing and order-of-operations control details using \yaq{} in the Supplementary Material.

@@

At the bottom, we graphically represent the integration between \yaq{} and the Bluesky project\cite{AllanDanielB2019a}.
Bluesky provides a powerful orchestration layer for conducting and recording data for a wide variety of experimental procedures.
Bluesky has no built-in hardware interface support, instead relying on packages to create hardware-interfaces that are compatible with their Hardware Protocol defintions.
\texttt{yaqc-bluesky}\footnote{\texttt{yaqc-bluesky} available on PyPI at https://pypi.org/project/yaqc-bluesky/} is a specialized client which adapts any \yaq{} protocol into Bluesky's Hardware Protocol.
Similar translation layers could be built for a variety of orchestration layers such as PyMoDAQ\cite{WeberSebastien2021a}, Instrumental\cite{Bogdanowicz2022}, or TRSpectrometer\cite{trspectrometer}.

@@

Figure \ref{fig:histogram} shows the distribution of unique lines of Python code written to implement each of the daemons in our current ecosystem.
While lines of code is an imperfect metric, we use it here to represent the amount of work required to create a daemon for a new hardware peripheral.
Most interfaces, such as Brooks MFC\footnote{\texttt{yaqd-brooks-mfc-gf} source code available on GitHub at \texttt{https://github.com/yaq-project/yaqd-brooks/}}, have been implemented in fewer than 100 lines of Python code.
Even the most complicated daemons are implemented in about 300 lines.
Implementing \yaq{} daemons using Python is enabled through our own tooling mentioned above and the large and growing ecosystem of hardware interface libraries that Python now provides\cite{pyserial, pyusb, pyvisa, pymodbus}.
In our experience, the process of creating a new daemon involves about a day of work after mastering communication with the hardware.
There are currently 72 daemons in the yaq project supporting at least 66 types of hardware, noting that some daemons support the same hardware and others are software only.
Because \yaq{} is standards based, anyone can design and publish new daemons extending our hardware support.
A living list of all daemons and supported hardware can be found on the \yaq{} website.\footnote{See https://yaq.fyi/daemons/ and https://yaq.fyi/hardware/ for living lists of all \yaq{} daemons and supported hardware, respectively.}

@@

Software built by scientists often has incomplete and inadequate documentation. \cite{SegalJudith2007a}
\yaq{} attempts to automate daemon documentation as much as possible.
Our website, https://yaq.fyi, automatically builds generated reference pages for all known protocols.
These pages are automatically updated when new versions are published.
Our website also contains written and video tutorials on \yaq{} usage and development.

We have designed \yaq{} to be easier to deploy and maintain when compared to EPICS and TANGO.
\yaq{} daemons are Python packages that can be installed on any platform using pip or conda.
There is no need for centralized management servers or databases when using \yaq{}.
The Avro RPC standard unambiguously describes the message signatures for any protocol, so users are always aware of the capabilities their hardware supports.
Traits enable interchangability where possible.
\yaq{} can easily be used alongside other experiment control software.
We provide tooling (\texttt{yaqd-control}) which makes it easy to configure, list, and manage daemons as background services on Windows, MacOS, or Linux (See SI for more details).

>>>>

Modern instrumentation development often involves incorporation of many dissimilar hardware peripherals into a single unified instrument.
Increasing availability of modular hardware has brought greater instrument complexity to small research groups.
This complexity stretches the capability of traditional, monolithic orchestration software.
In many cases, a lack of software flexibility leads creative researchers to feel frustrated, unable to perform experiments they envision.
Herein we describe \yaq{}, a software project defining a new standardized way of communicating with diverse hardware peripherals.
\yaq{} encourages a highly modular approach to experimental software development which is well suited to address the experimental flexibility needs of complex instruments.
\yaq{} is designed to overcome hardware communication barriers that are insurmountable with typical experimental software.
A large number of hardware peripherals are already supported, with tooling available to expand support.
The \yaq{} standard enables collaboration among multiple research groups, increasing code quality while lowering development effort.

@@

We have created a new network-based communication standard for scientific instrumentation, \yaq{}.
This standard borrows the most important ideas from established projects used by large user facilities while retaining the simplicity appropriate for small research labs.
We have built this standard to be self-describing, portable, and reusable wherever possible.
Our primary goal has been to make an interface which simplifies orchestration software development as much as possible.

@@

In \yaq{}, communication between daemons and clients is performed over TCP/IP using Apache Avro RPC \cite{AvroSpecification}.
Avro provides an agreed upon standard for efficient serialization of data and method calls from a remote (client) process.
Practically, the \yaq{} interface looks like a collection of methods or functions, which Avro calls ``messages''.
Each message has defined input parameters and output return types.
A sensor might implement a message called ``\texttt{get\_measured}'' which takes no parameters and returns a dictionary mapping channel names to numeric or array data.
A motor would implement a pair of messages for setting and reading back the motor position: ``\texttt{set\_position(float position)} $\rightarrow$ \texttt{null}'' and ``\texttt{get\_position()} $\rightarrow$ \texttt{float}''.
Messages make up the lowest level functionality of the \yaq{} interface.
Each individual communication between client and daemon involves one message being requested by the client and the response returned from the daemon.
Each daemon supports a collection of messages for its unique functionality, called the ``protocol''.
When a client first connects to a daemon over TCP/IP, the daemon provides a complete description of its own protocol.
We therefore refer to the daemons as ``self-describing''.

\yaq{} introduces a concept called ``traits'' which are collections of related messages that are shared among multiple protocols.
Motors implement the ``\texttt{has-position}'' trait, which defines ``\texttt{set\_position}'', ``\texttt{get\_position}'', and ``\texttt{get\_units}''.
Sensors would implement the ``\texttt{is-sensor}'' trait which defines ``\texttt{get\_measured}'', ``\texttt{get\_channel\_names}'', and ``\texttt{get\_channel\_units}''.
Protocols which implement a trait must support all of the messages from the trait.
Importantly, specific protocols can also implement arbitrary additional messages that are not defined by any trait.

The first hardware interface barrier: Multiple incompatible interfaces are used to communicate with each component of the system. % consider making this sentence (and the next paragraph starts) subsection headings
\yaq{} provides a unified TCP/IP interface to all hardware peripherals based on the well-described Avro RPC protocol.
The trait system was introduced in pursuit of our primary goal of easing the client development process.
Clients can trust that protocols that implement a given trait will behave in similar ways.
The standardized \yaq{} interface presents the same set of interactions for client-side scientific code, simplifying the experience of using hardware.

@@

The third hardware interface barrier: Some hardware interfaces are blocking.
Monolithic orchestration software often necessitates separate threads for each component hardware interface, a fragile pattern in which small mistakes become both critical and elusive errors.
In \yaq{}, each daemon is only responsible for a single hardware interface.
%Each \yaq{} daemon runs as a separate process (possibly not even on the same machine) and thus many of the hard concurrency problems with communicating with hardware are mitigated.
It is expected that each message call over the \yaq{} interface will return rapidly, ensuring that client applications are not blocked for extended periods of time.
This principle applies even when the message starts an action that might take several seconds to complete, such as homing a motor or initiating a measurement for a sensor.
In these instances, the initial message simply starts the action and returns, with a separate message provided to retrieve results when they exist.
In order to know how long to wait, the ``is-daemon`` trait provides a message called ``is\_busy'', which returns "true" while the long running action is not complete, and "false" once it is finished.
Additionally, multiple clients can communicate with the same daemon simultaneously.
A complex instrument may involve multiple operators watching sensor data in real time, while one program is orchestrating the hardware and recording the data.

@@

At the bottom, we show several features associated with the integration between \yaq{} and the Bluesky project\cite{AllanDanielB2019a}.
Bluesky provides a powerful orchestration layer for conducting and recording data for a wide variety of experimental procedures.
Shown in Figure \ref{fig:foundation}, we see one such GUI for parameterizing a Bluesky procedure and a representation of the resultant data.
\texttt{yaqc-bluesky}\footnote{\texttt{yaqc-bluesky} available on PyPI at https://pypi.org/project/yaqc-bluesky/} provides a bridge to the Bluesky ecosystem.
Similar translation layers could be built for a variety of orchestration layers such as PyMoDAQ\cite{WeberSebastien2021a}, Instrumental\cite{Bogdanowicz2022}, or TRSpectrometer\cite{trspectrometer}.

@@

Figure \ref{fig:histogram} shows the distribution of unique lines of Python code written to implement each of the daemons in our current ecosystem.
While lines of code is an imperfect metric, we use it here to represent the amount of work required to create a daemon for a new hardware peripheral.
Most interfaces, such as Brooks MFC\footnote{\texttt{yaqd-brooks-mfc-gf} source code available on GitHub at \texttt{https://github.com/yaq-project/yaqd-brooks/}}, have been implemented in fewer than 100 lines of Python code.
Even the most complicated daemons are implemented in about 300 lines.
Implementing \yaq{} daemons using Python is enabled through our own tooling mentioned above and the large and growing ecosystem of hardware interface tooling libraries that Python now provides\cite{pyserial, pyusb, pyvisa, pymodbus}.
In our experience, the process of creating a new daemon involves about a day of work after mastering communication with the hardware.
There are currently 72 daemons in the yaq project supporting at least 66 types of hardware, noting that some daemons support the same hardware and others are software only.
Because \yaq{} is standards based, anyone can design and publish new daemons extending our hardware support.
A living list of all daemons and supported hardware can be found on the \yaq{} website.\footnote{See https://yaq.fyi/daemons/ and https://yaq.fyi/hardware/ for living lists of all \yaq{} daemons and supported hardware, respectively.}

@@

Software documentation is famously difficult and thankless work.
\yaq{} attempts to automate daemon documentation as much as possible.
Our website, https://yaq.fyi, automatically builds generated reference pages for all known protocols.
These pages are automatically updated when new versions are published.
Our website also contains written and video tutorials on \yaq{} usage and development.
